---
title: "02a_date_extrapolation"
author: "Julia Pop"
date: "2025-10-06"
output: html_document
---
```{r setup, include=FALSE}
# code to troublesheet file path issues. 
# By default, knitr uses the folder containing the .Rmd file as
# the working directory. This code allows us to set the folder
# with our .rProj in it as the root. 

# find_root looks upwards from the .Rmd until it finds a file or
# folder that matches the criteria "folder with a .Rproj file in
# it."

knitr::opts_knit$set(root.dir = rprojroot::find_root(rprojroot::is_rstudio_project))
```

#Libraries -- add these here
```{r}
source("scripts/libraries.R")
```

# Pull in data
```{r}
all_lakes <- read.csv("output/all_lakes.csv")

UFM_short <- read.csv("output/UFM_short.csv") %>%
  select(-X)
UFM_wide <- read.csv("output/UFM_wide.csv") %>%
  select(-X)
LFM_short <- read.csv("output/LFM_short.csv") %>%
  select(-X)
TRK_short <- read.csv("output/TRK_short.csv") %>%
  select(-X)

# read in bulk density data
bulk_density <- read_excel("data/bulk_density_export_from_sent_to_flett_20250602_saved_20250909.xlsx") %>%
  rename(bd = `bulk_density_g/cm3`)

# read in UFM metadata
UFM_metadata <- read_excel("data/0_UFM1_core_metadata_20250726.xlsx") %>% 
  select(lake_name:depth_bottom) %>%
  left_join(bulk_density %>% 
              select(sample_ID, bd),
            by = "sample_ID")
```

# Age-depth models
```{r}
# read in raw info from Flett for reference
all_dates <- read_excel("data/Flett_dates_simplified_20250917.xlsx")

# adms from 00_data_input_and...
UFM_adm <- readRDS("output/UFM_adm.rds")
LFM_adm <- readRDS("output/LFM_adm.rds")
TRK_adm <- readRDS("output/TRK_adm.rds")
# now we have an adm to mess with!
```


# Date extrapolation
##TRK
```{r}
# see what predict did to the sedimentation rates and if I should set my own age depth model... (at an avg sed rate for example)
names(TRK_short)
# plot the extrapolated adm...
plot(depth_plot ~ year_plot_extrap, data = TRK_short, type = "l", 
     xlab = "Year CE", ylab = "Depth (cm)")
points(TRK_short$year_plot_extrap, TRK_short$depth_plot, pch = 20, cex = 0.3)
abline(v = 1982.209, lty = "aa")

# plot the interpolated adm
plot(TRK_adm)

```

##UFM
```{r}
# dates_clean length (lowest depth_bottom)
  # UFM: 5 cm
  # LFM: 18 cm
  # TRK: 11 cm

  # UFM_adm is only for the interpolated dates. 
  # making another adm here for the extrapolated dates.

# NOTE: SWITCHING HERE FROM EXTRAP BOTTOM TO REGULAR depth_bottom

  # pull in extrapolated estimate (Flett's file with an approximation 
  # on the timing of UFM1_024)

  UFM_estimate <- read_xlsx("data/UFM_estimate.xlsx")

  # bind the extrapolated estimate to the metadata, cut out nas
  UFM_dates_try <- UFM_estimate %>%
    left_join(UFM_metadata %>% select(-depth_top, -depth_bottom),
              by = "sample_ID") %>%
    relocate(all_of(c(names(UFM_metadata)))) %>%
    mutate(year_CE = 2025 - extrap_bottom_age_lin) %>%
    filter(!is.na(extrap_bottom_age_lin))
  
  # adjust the old adm
    UFM_adm <- age_depth_model(UFM_dates_try,
                               depth = extrap_bottom,
                               age = year_CE)

  # goes back pretty far but this looks good. age:depth rate is constant.
  plot(UFM_adm)

  # extrapolate the dates to 44 cm using the predict function. 
  UFM_extrap <- predict(UFM_adm, depth = seq(0.25, 44, 0.25)) %>%
    mutate(year_CE = age) %>%
    select(depth, year_CE)

## create new columns in the interp df based on the ending digits of depth
## to easily join the interpolated ages later

  # functions to categorize depths by ending digits
  is_whole  <- function(x, tol = 1e-8) abs(x %% 1) < tol
  is_half   <- function(x, tol = 1e-8) abs(x %% 1 - 0.5) < tol
  is_three_quarters <- function(x, tol = 1e-8) abs(x %% 1 - 0.75) < tol
  
  # make new columns in the extrapolating adm df
  UFM_extrap <- UFM_extrap %>%
    mutate(
      depth_top    = ifelse(is_half(depth), depth, NA_real_),
      depth_plot   = ifelse(is_three_quarters(depth), depth, NA_real_),
      depth_bottom = ifelse(is_whole(depth), depth, NA_real_)
    ) %>%
    select(depth, year_CE, depth_top, depth_plot, depth_bottom)
  
# Join new columns to UFM_short: top, plot, then bottom
  
UFM_short_extrap <- UFM_short %>%
  left_join(
    UFM_extrap %>% 
      select(depth_top, year_CE) %>%
      rename(year_top_extrap = year_CE),
    by = "depth_top")

UFM_short_extrap <- UFM_short_extrap %>%
  left_join(
    UFM_extrap %>% 
      select(depth_plot, year_CE) %>%
      rename(year_plot_extrap = year_CE),
    by = "depth_plot")

UFM_short_extrap <- UFM_short_extrap %>%
  left_join(
    UFM_extrap %>% 
      select(depth_bottom, year_CE) %>%
      rename(year_bottom_extrap = year_CE),
    by = "depth_bottom") 

# Repeat for UFM_wide
UFM_wide <- UFM_wide %>%
  left_join(
    UFM_short_extrap %>%
      ungroup() %>% # keep r from adding these variables twice
      select(sample_ID, year_top_extrap:year_bottom_extrap), 
    by = "sample_ID"
  )
```

##LFM
```{r}
# We have 43 cm of LFM2. According to our adm how many years should that go back? 
predict(LFM_adm, depth = 43)$age # to ~1613

# next steps: 
# extrapolate the dates to 43 cm using the predict function. 
# calculate sedimentation rates using the code from 00_analysis
# see what predict did to the sedimentation rates and if I should set my own age depth model... (at an avg sed rate for example)

# extrapolate the dates to 43 cm using the predict function. 
LFM_extrap <- predict(LFM_adm, depth = seq(0.25, 43, 0.25)) %>%
  mutate(year_CE = age) %>%
  select(depth, year_CE)

# plot the extrapolated adm...
plot(depth ~ year_CE, data = LFM_extrap, type = "l", 
     xlab = "Year CE", ylab = "Depth (cm)")
points(LFM_extrap$year_CE, LFM_extrap$depth, pch = 20, cex = 0.3)
abline(v = 1861.67, lty = "aa")

# plot the interpolated adm
plot(LFM_adm)

```

###Calculate depth top, bottom, plot, duration and weight
```{r}
## create new columns in the extrap df based on the ending digits of depth
## to easily join the interpolated ages later

  # functions to categorize depths by ending digits
  is_whole  <- function(x, tol = 1e-8) abs(x %% 1) < tol
  is_half   <- function(x, tol = 1e-8) abs(x %% 1 - 0.5) < tol
  is_three_quarters <- function(x, tol = 1e-8) abs(x %% 1 - 0.75) < tol
  
  # making new columns
  LFM_extrap <- LFM_extrap %>%
    mutate(
      depth_top    = ifelse(is_half(depth), depth, NA_real_),
      depth_plot   = ifelse(is_three_quarters(depth), depth, NA_real_),
      depth_bottom = ifelse(is_whole(depth), depth, NA_real_)
    ) %>%
    select(depth, year_CE, depth_top, depth_plot, depth_bottom)

# Join new columns to LFM_wide: top, plot, then bottom
LFM_short_extrap <- LFM_short %>%
  left_join(
    LFM_extrap %>% 
      select(depth_top, year_CE) %>%
      rename(year_top_extrap = year_CE),
    by = "depth_top")

LFM_short_extrap <- LFM_short_extrap %>%
  left_join(
    LFM_extrap %>% 
      select(depth_plot, year_CE) %>%
      rename(year_plot_extrap = year_CE),
    by = "depth_plot")

LFM_short_extrap <- LFM_short_extrap %>%
  left_join(
    LFM_extrap %>% 
      select(depth_bottom, year_CE) %>%
      rename(year_bottom_extrap = year_CE),
    by = "depth_bottom") %>%
  mutate(duration_extrap = year_top_extrap - year_bottom_extrap) # calculate duration 

duration_mean_LFM_short_extrap <- mean(LFM_short_extrap$duration_extrap, na.rm = TRUE)
LFM_short_extrap <- mutate(LFM_short_extrap, 
                           weight_extrap = duration_extrap / duration_mean_LFM_short_extrap)

# ages stay the same, but the weight term changes because the average is pulled off by the extrapolated dates... don't use it in the final analysis? tbd. at least can say I looked into it. 

# specify which ages were interpolated and which were extrapolated
LFM_short_extrap <- LFM_short_extrap %>%
  mutate(age_cert = ifelse(year_plot_extrap > 1861, "interp", "extrap"))
```

###See what changed:
```{r}

LFM_short_extrap_view <- LFM_short_extrap %>%
       select(lake_name, 
              sample_ID, 
              sample_range, 
              depth_top, 
              depth_bottom, 
              depth_plot, 
              year_CE_top, 
              year_CE_plot, 
              year_CE_bottom, 
              duration, 
              weight, 
              s,
              year_top_extrap,
              year_plot_extrap, 
              year_bottom_extrap,
              duration_extrap,
              weight_extrap)

View(LFM_short_extrap_view)

rm(LFM_short_extrap_view)

```

Okay. If we want to model gams over the whole core using extrapolated dates (and only interpret them in the interpolated sections) we might need to drop the weight term (bc it's pulled off by the extrapolated dates) and all the accumulation rate metrics... 

Whether it's reasonable to interpolate so far back is another question. Can show the graph from above...

UPDATE: not going to model the whole core, just going to plot the data and model the dated section. 

**Or run it the regular way first then see if running first derivs on fluxes gives any significant periods of change?**

### Removing variables from LFM_wide...
```{r}
names(LFM_short_extrap)

# LFM_short_extrap <- LFM_short_extrap %>%
#   select(
#    -"X",
#    -"C.mar",
#    -"N.mar",
#    -"year_CE_top",
#    -"year_CE_plot",
#    -"year_CE_bottom",
#    -"duration",
#    -"weight",
#    -"s",
#    -"duration_extrap", 
#    -"weight_extrap", 
#    -"fuco_flux",
#    -"allo_flux",
#    -"diato_flux",
#    -"lutein_flux",
#    -"cantha_flux",
#    -"chl_b_flux",
#    -"chl_a_flux",
#    -"chl_ap_flux",
#    -"echine_flux", 
#    -"phaeo_b_flux",
#    -"pheo_a_flux",
#    -"b_car_flux"
#  )
```


#Export
```{r}
# export important dfs as csvs 
# LFM_wide won't have the extrapolated dates. TRK_wide does. 
write.csv(LFM_short_extrap, file="output/LFM_short_extrap.csv")

write.csv(UFM_short_extrap, file="output/UFM_short_extrap.csv")
write.csv(UFM_wide, file="output/UFM_wide.csv")
```
